Preliminary insights into the use of pottery and culinary pratices at Guijiabao site in southwestern China

### 2024-01-21 analysis/paper/paper.Rmd
substantial edits

---
title: "Preliminary insights into the use of pottery and culinary pratices at Guijiabao site in southwestern China"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or terrestrial animals across areas in southwestern China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region. It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and millets came even earlier. However, to what extent new crops were used by local peoples at that time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwestern China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis. The results show signals of terrestrial animals in most potsherds and a C~3~ plant source in one potshard, during the Neolithic period. Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming. This study also demonstrates an approach using organic residue analysis to explore the extent of the adoption of farming during the Neolithic period in southwestern China.

keywords: |
  Southwestern China; Neolithic period; Lipid residue analysis; Stable isotope analysis; Ceramics; Culinary practices
highlights: |
  These are the highlights. 
---

@@

# Introduction
<!-- intro to the topic and research aim-->

The movement of Neolithic populations has long been discussed in archaeology to understand the processes of agricultural diffusion, regional variations, and their consequences [@bellwood2023; @Davison2006]. In southwest China, the emergence of farming practices was thought to be associated with the migrating population from central or northern China in the Neolithic period [@Deng2019]. Crop remains, including foxtail millet, broomcorn millet, and rice, found in prehistoric southwest China that can be traced back to as early as 4700 BP in Sichuan and around 4600 BP in western Yunnan [@Jiang2016; @Deng2019; @YunnanMuseum1981; @Zhang2010; @Dal2022]. Despite different views on the routes of agricultural dispersal [@Guedes2013; @Guedes2018; @Liu2022], it is generally believed that mixed farming was the main economy in southwest China after the introduction of agriculture [@Deng2019]. However, more studies still needed to understand how migrant communities interacted with new environments, especially in an area with high diversity of landscape, such as high mountain, plateau, basin, and terrace.

This study focuses mainly on archaeological sites located in the Hengduan Mountain ranges, a region in southwest China connecting the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south. As a critical passage for population movement, the Hengduan Mountain ranges and valleys facilitate not only material exchange [@Tong1986; @Hein2014; @Hein2016; @Ma2021], but also the spread of agriculture even further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017]. Recent excavations at the Guijiabao site in Yanyuan Basin, the middle Hengduan Mountain ranges, is a key site with the earliest Neolithic human occupation in this region dating to 5000 B.P., continuing to the historical period [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022]. The complete stratigraphy at Guijiabao provides a unique opportunity to investigate questions relative to the subsistence since the Neolithic period in a basin surrounded by high mountains. For example, what is the role of agriculture in local environments? What other resources were obtained in addition to crops? In this study, we present our preliminary analyses on archaeological potsherds from the Guijiabao site and the Gaoshan site in southwest China using organic residue analysis of lipids.

Although the remains of animal bones and plants offer evidence for subsistence, their preservation heavily depend on environmental conditions. Geographical regions with seasonal heavy precipitation and acidic sediments, such as Southwest Sichuan and the middle Hunduan Mountain [@Li1989], tend to lead to poor preservation of archaeological remains. In such case, organic residue analysis is a powerful tool for studying ancient food practices by directly examining residues in pottery. Organic residues, specifically lipids, can be well preserved in the porous fabric of pottery for over thousand years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017; @Dudd1999; @Miller2020]. The growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies [@Kwak2020], and culinary practices across a region over time [@Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].

Through identification of lipid biomarkers and compound-specific carbon isotope analysis based on geochemical techniques, archaeologists are able to distinguish food origins between plant oils, ruminant fats, diary products, and aquatic/marine resources [@Copley2005; @Craig2007; @Evershed1994]. Recent studies further successfully identified different types of plants processed in archaeological vessels, such as broomcorn millet [@Bossard2013; @Heron2016; @Ganzarolli2018], starchy plants [@Shoda2018], and cereals [@Hammann2018; @Colonese2017]. A growing case studies for ceramics in China present the robustness of organic residue analysis for understanding regional cooking practices [@Shoda2018; @Lyu2024; @Han2022; @Lyu2024], distinct subsistence strategies between low- and high altitude zones [@Zhang2022], and the impact of nomadic pastoralism on northwest China [@Sun2023]. In this study, we attempt to investigate pottery use and culinary practices in the middle Hengduan mountains to understand the local subsistence by focusing on the ceramics from the Guijiabao and Gaoshan sites.

@@

<!--Yanyuan area and Guijiabao site-->
The Guijiabao site is located in the southwest of Yanyuan Basin, a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)), with an elevation of 2400 meters. The first survey at Guijiabao in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean and ground houses, post holes, ash pits, ditches, and burials, suggesting a settlement site [@Hao2022; @Guijiabao2017a]. The site was occupied since the Neolithic period (5000-3700 BP), continued to the Bronze Age (3200-2700 BP) and the late North Song to early Ming dynasties (1200-500 BP), based on 19 radiocarbon dates [@Hao2022]. The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the artifact assemblages and ceramic styles

<!--the major finding at GJB-->
Ceramics is the major finding at Guijiabao. The common vessel forms include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a]. The evidence of chars on their surfaces suggest that they might be used for cooking [@Guijiabao2017a; @Hao2022]. This research also attempts to confirm that these pottery vessels were used as cooking/storage wares. Some ceramic decorations present northern styles, while others with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of close relationships or interaction between Guijiabao and neighboring areas, the northern and southern parts of the Hengduan Mountains [@Guijiabao2017a]. The early excavation reveals the presence of cultivated crops at Guijiabao around 4700 BP, including foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*), earlier than previously thought in Yanyuan [@Guijiabao2016]. In later excavations, rice grains (*Oryza sativa*) were also found throughout the Neolithic period, but in a smaller number than millet [@Hao2022; @Huan2022].

@@

Although mixed agriculture of millet and rice observed in southwest China, it seems there are variations in crop cultivation across the region. The macro-botanical and phytolith analyses of the Guijiabao site show that millet was the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, botanical evidence from most sites in the Chengdu plain reveal a predominance of rice, such as at the Gaoshan and Baodun sites [@JiangGaoshan2017], and the Baiyangcun site in Yunnan [@Jiang2016; @Henglanshan2014]. Moreover, most research focuses on the agricultural practices, what other resources utilized by early farmer in new environments remains unclear. Here we seek to examine direct evidence from pottery residues to provide clues on culinary practices through case study in southwest China using geochemical techniques.

@@

## Sampling
As a preliminary research on food residues preserved in ceramic vessels, we sampled pottery fragments from cultural layers at the residential area of Guijiabao (n = 5) to test the feasibility of organic geochemical approaches (Table \@ref(tab:sherds-table)). As a comparison, we sampled one fragment from Goashan. Those pottery fragments were identified as typical types with well-defined contexts, as indicated by radiocarbon dates ranging from the Neolithic period (ca. 4500 BP) to the historical period (ca. 1000 BP). The pottery samples are mostly base fragments and fine sand-tempered, with a few displaying impressed decorations. Through examining potsherds from early to late period at Guijiabao, we seek to explore the potential of organic geochemical approach across different time period.

@@

## Lipid extraction
The surface of each pottery sample was removed to prevent external contamination from environments and handling. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018]. To obtain a higher yield, free fatty acids were recovered from bound lipids through saponification, conducted by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heated at 70 °C overnight [@Kaluzna2016; @Craig2004]. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of the extract was treated with 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) containing 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene for gas chromatography mass spectrometry (GC-MS).

The remaining 95% aliquot was fractioned using solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL). About 5% aliquot was analyzed by gas chromatography with flame-ionization detection (GC-FID) for quantification, and the remaining sample was analyzed using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS) [@Craig2007; @Evershed1994]. Analytical blanks were prepared and analyzed using the same procedure as the pottery samples to monitor potential contamination sources from solvents and reagents. Internal standards, FA C~21:0~ and 5-α Cholestane or n-tetratriacontane, were added before and after the lipid extraction.

## GC methods
The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector (Agilent technologies, USA). Samples were injected in splitless mode at 320 °C. Compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness; J&W Scientific, USA). The carrier gas was helium flowing at a rate of 1.3 ml/min. The oven temperature was set at 90 °C for 1 minute, then increased at a rate of 5 °C/min to 320 °C and held for 23 minutes. In addition to scan mode, a selected ion monitoring (SIM) mode was used to further detect miliacin (olean-18-en-3β-ol methyl ether; *m/z* 189, 204, 231, 425, 440) [@Heron2016]. The oven temperature was set at 50 °C for 1 min, then raised at a rate of 10°C/min to 310 °C and held for 15 min. A authentic standard of miliacin was also ran to monitor the retention time.

The GC-FID analysis of FAMEs was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode (Agilent technologies, USA). The carrier gas was helium flowing at a rate of 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was set at 90°C for 1.3 minute, then raised at a rate of 5°C/min to 320°C and held for 22 minutes. The GC-C-IRMS analyses of FAMEs of palmitic (C~16:0~) and stearic (C~18:0~) were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface (Thermo Fisher Scientific, USA) in splitless mode. Separation was carried out using the same model of column for GC-MS. Helium was used as a gas carrier with a constant flow rate of 1.24 ml/minute. The oven temperature was set at 90 °C for 1.5 minutes, then raised at a rate of 5 °C/min to 320 °C and held for 18 minutes.

@@

# add sample name
meth_all <-
  meths %>%
  mutate(sample = paste0(parse_number(sample),
                         str_extract(sample, pattern = "_00[0-9]"))) %>%
  mutate(name = case_when(
    sample %in% c("210729_001") ~ "CDG-062",
    sample %in% c("210729_002") ~ "SYG-3",
    sample %in% c("210729_003") ~ "SYG-TN13-E22-2#1",
    sample %in% c("210729_004") ~ "SYG-TN13-E22-1",
    sample %in% c("210729_005") ~ "SYG-TN13-E22-2#2",
    sample %in% c("210729_006") ~ "SYG-TN13-E23-3")) %>%
  mutate(name = factor(name, levels = c("SYG-TN13-E22-1",
                                        "SYG-TN13-E22-2#1",
                                        "SYG-TN13-E22-2#2",
                                        "SYG-TN13-E23-3",
                                        "SYG-3",
                                        "CDG-062")))
```

```{r highlight-compounds, cache=TRUE}
# highlight target compounds
highlight_common <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 21.6 ~ "C[14:0]",
    retention == 23.6 ~ "C[15:0]",
    retention == 25.5 ~ "C[16:0]",
    retention == 27.3 ~ "C[17:0]",
    retention == 28.4 ~ "C[18:1]",
    retention == 29.1 ~ "C[18:0]",
    retention == 32.3 ~ "C[20:0]",
    retention == 35.4 ~ "C[22:0]")) %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
  filter(File > 100000)

highlight_IS <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 33.9 ~ "IS",
    retention == 41.3 ~ "IS")) %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062"))

# highlight the compounds in the potsherd with high concentration
highlight_SYG2_2 <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 21.6 ~ "C[14:0]",
    retention == 25.5 ~ "C[16:0]",
    retention == 27.3 ~ "C[17:0]",
    retention == 28.4 ~ "C[18:1]",
    retention == 29.1 ~ "C[18:0]",
    retention == 31.0 ~ "C[19:0]",
    retention == 32.4 ~ "C[20:0]",
    retention == 35.4 ~ "C[22:0]",
    retention == 36.8 ~ "C[23:0]",
    retention == 38.2 ~ "C[24:0]",
    retention == 40.9 ~ "C[26:0]",
    retention == 33.9 ~ "IS",
    retention == 41.3 ~ "IS",
    retention == 45.6 ~ "C[30:0]",
    retention == 47.8 ~ "C[32:0]")) %>%
  filter(name == "SYG-TN13-E22-2#2")
```

>>>>

---
title: "Preliminary insights into the use of pottery and culinary pratices at Guijiabao site in southwestern China"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |

     The emergence of mixed agriculture of millet and rice in Neolithic southwestern China has been a topic of extensive discussion. Despite the recovery of both millet and rice grains, the varying proportions observed across the area underscore the complexity of the process of agricutural dispersal. Further investigation is needed to understand how new crops were integrated into southwestern China, characterized by diverse environmental conditions and landscapes. This study contributes to this understanding by exploring subsistence strategies in southwestern Sichuan through a preliminary analysis of lipid residues in pottery from the Guijiabao site using organic geochemical analyses. Situated in the middle Hengduan mountain ranges, the Guijiabao site presents well-preserved stratigraphic records spanning from the Neolithic to the historical period, providing valuable insights into local subsistence practices in a highland environment. Results reveal the detection of ruminant adipose fats in the majority of potsherds, with a few showing plant inputs such as broomcorn millet. This suggests that animals played a substantial role in the local diet, alongside the consumption of millet. These culinary practices also indicate that hunting may have remained a significant subsistence activity in the high-altitude region of southwestern China, even after the introduction of crops.

keywords: |
  Southwestern China; Neolithic period; Lipid residue analysis; Compound-specific carbon isotope analysis; Pottery; Culinary practices
highlights: |
  These are the highlights. 
---

@@

# Introduction

<!-- intro to the topic and research aim-->
The movement of Neolithic populations has long been discussed in archaeology to understand the processes of agricultural diffusion, regional variations, and their consequences [@Bellwood2023; @Davison2006]. In southwest China, the emergence of farming practices was thought to be associated with the migrating population from central or northern China in the Neolithic period [@Deng2019]. Crop remains found in prehistoric southwest China, including foxtail millet, broomcorn millet, and rice, can be traced back to as early as 5000 BP in southwest Sichuan and around 4600 BP in western Yunnan [@Jiang2016; @Deng2019; @YunnanMuseum1981; @Zhang2010; @Dal2022]. Despite different views on the routes of agricultural dispersal [@Guedes2013; @Guedes2018; @Liu2022; @Huan2022], it is generally believed that mixed farming was the main agricultural practice in southwest China [@Deng2019]. However, more studies are still needed to understand how migrant communities interacted with new environments, especially in an area with a high diversity of landscapes, such as high mountains, plateaus, basins, and terraces.

<!--Guijaibao as a research area-->
This study focuses mainly on archaeological sites located in the Hengduan Mountain ranges, a region in southwest China connecting the Tibetan Plateau to the northwest, the Sichuan Basin to the east, and the Yunnan Plateau to the south. As a critical passage for population movement, the Hengduan Mountain ranges and valleys facilitate not only material exchange [@Tong1986; @Hein2014; @Hein2016; @Ma2021], but also the spread of agriculture even further to mainland Southeast Asia since the Neolithic period [@Liu2022; @Stevens2017]. The Guijiabao site in Yanyuan Basin, the middle Hengduan Mountain ranges, are a key site with the earliest Neolithic human occupation in this region dating to 5000 B.P., continuing to the historical period [@Guijiabao2017a; @Guijiabao2017b; @Huan2022; @Hao2022]. The complete stratigraphy at Guijiabao provides a unique opportunity to investigate questions relative to subsistence since the Neolithic period in a basin surrounded by high mountains. For example, what is the role of agriculture in local environments? What other resources were obtained in addition to crops? In this study, we present our preliminary analyses of archaeological potsherds from the Guijiabao site and the Gaoshan site in southwest China using organic residue analysis of lipids.

<!--justifying organic residue analysis-->
Although the remains of animal bones and plants offer evidence of subsistence, their preservation heavily depends on environmental conditions. Geographical regions with seasonal heavy precipitation and acidic sediments, such as Southwest Sichuan and the middle Hunduan Mountain [@Li1989], tend to lead to poor preservation of archaeological remains. In such cases, organic residue analysis is a powerful tool for studying ancient food practices by directly examining residues in pottery. Organic residues, specifically lipids, can be well preserved in the porous fabric of pottery for thousands of years due to the protection of clay matrices from microbiological degradation and water leaching [@Evershed1993; @Evershed2008; @Roffet2017; @Dudd1999; @Miller2020]. The growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies [@Kwak2020], and culinary practices across a region over time [@Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020].

<!--regional application-->
By identifying lipid compositions and conducting compound-specific carbon isotope analysis, archaeologists are able to distinguish between plant oils, ruminant fats, dairy products, and aquatic/marine resources [@Copley2005; @Craig2007; @Evershed1994]. Recent studies have successfully identified various plant types processed in archaeological vessels, including broomcorn millet [@Bossard2013; @Heron2016; @Ganzarolli2018], starchy plants [@Shoda2018], and cereals [@Hammann2018; @Colonese2017]. The increasing use of ceramics in China highlights the robustness of organic residue analysis for comprehending regional cooking practices [@Shoda2018; @Lyu2024; @Han2022], discerning distinct subsistence strategies between low- and high-altitude zones [@Zhang2022], and exploring the impact of nomadic pastoralism on northwest China [@Sun2023]. In this study, we aim to investigate pottery use and culinary practices in the middle Hengduan mountains, focusing on ceramics from the Guijiabao and Gaoshan sites to understand local subsistence.

@@

<!--Yanyuan area and Guijiabao site-->
The Guijiabao site is located in the southwest of Yanyuan Basin, a pull-apart basin in the middle eastern part of the Hengduan Mountain ranges (Figure \@ref(fig:SW-china-map)), with an elevation of 2400 meters. The first survey at Guijiabao in 2015 and subsequent excavations from 2016 to 2019 revealed abundant archaeological materials and features, such as semi-subterranean and ground houses, post holes, ash pits, ditches, and burials, suggesting a settlement site [@Hao2022; @Guijiabao2017a]. The site has been occupied since the Neolithic period (5000-3700 BP), continued to the Bronze Age (3200-2700 BP) and the historical period (1200-500 BP) corresponding to the late North Song to early Ming dynasties, based on radiocarbon dates [@Hao2022; @Huan2022]. The Neolithic period can be further divided into two subphases, 5000-4500 BP and 4500-3700 BP, according to the artifact assemblages and ceramic styles.

<!--the major finding at GJB-->
Ceramics are the major finding at Guijiabao. The common vessel forms include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Ceramics with charred surfaces suggest that they might be used for cooking. This research also attempts to confirm that these pottery vessels were used as cooking or storage wares. Some ceramic decorations present northern styles, while others with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of close relationships or interaction between Guijiabao and neighboring areas, the northern and southern parts of the Hengduan Mountains [@Guijiabao2017a]. The flotation analysis of soil samples from the 2015 excavation demonstrates the predominance of cultivated broomcorn millet (*Panicum miliaceum*) and foxtail millet (*Setaria italica*) at Guijiabao during the Neolithic period [@Guijiabao2016]. Rice grains (*Oryza sativa*) were also identified in later excavations from the Neolithic period based on macro botanical evidence and phytolith analysis [@Hao2022; @Huan2022], though in a smaller number than millet.

@@

While mixed agriculture of millet and rice was practiced in southwestern China, variations in their proportions across the region have been observed. Macro-botanical and phytolith analyses of the Guijiabao site indicate that millet was the primary crop, aligning with findings at the Yingpanshan site in Sichuan [@Zhao2011]. Conversely, botanical evidence from numerous sites in the Chengdu plain suggests a prevalence of rice, as seen at the Gaoshan and Baodun sites [@JiangGaoshan2017], and the Baiyangcun site in Yunnan [@Jiang2016; @Henglanshan2014]. Moreover, while most research focuses on agricultural practices, there is a lack of clarity regarding other resources utilized by early farmers in new environments. This study aims to address this gap by examining direct evidence from pottery residues, using geochemical techniques to provide insights into culinary practices through a case study in southwest China.

@@

## Sampling

As a preliminary study to assess the feasibility of employing organic geochemical approaches, we collected five pottery fragments from cultural layers in the residential area of Guijiabao (sample code starting with SYG) (Table \@ref(tab:sherds-table)). For comparison, one fragment was sampled from Goashan (sample code starting with CDG). Those pottery fragments represent typical types with well-defined contexts, with their radiocarbon dates ranging from the Neolithic period (ca. 4500 BP) to the historical period (ca. 1000 BP). The pottery samples are mostly base fragments and fine sand-tempered, with a few displaying impressed decorations. By examining potsherds from the early and late periods at Guijiabao, we aim to explore pottery use across time periods and the potential of utilizing an organic geochemical approach.

@@

## Lipid extraction

The surface of each pottery sample was removed to prevent external contamination from the environment and handling. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018]. To obtain a higher yield, free fatty acids were recovered from bound lipids through saponification, which was conducted by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating at 70 °C overnight [@Kaluzna2016; @Craig2004]. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of the extract was treated with 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) containing 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene for gas chromatography mass spectrometry (GC-MS).

The remaining 95% aliquot was fractioned using solid phase extraction (Aminopropyl, NH2, column with 0.5 g sorbent) with 8 mL dichloromethane:Isopropyl alcohol (3:1), 8 mL 4% acetic acid in ethyl ether, and 6 mL methanol. The second fraction was methylated to produce fatty acid methyl esters (FAMEs) with 1 mL dry hexane and 2 mL 10% (v/v) acetyl chloride in dry methanol, and heated at 70 °C overnight. The FAME derivatives were extracted with hexane ($3\times10$ mL). About a 5% aliquot was analyzed by gas chromatography with flame-ionization detection (GC-FID) for quantification, and the remaining sample was analyzed using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS) [@Craig2007; @Evershed1994]. Analytical blanks were prepared and analyzed using the same procedure as the pottery samples to monitor potential contamination sources from solvents and reagents. Internal standards, FA C~21:0~ and 5-α Cholestane or n-tetratriacontane, were added before and after the lipid extraction.

## GC methods
The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector (Agilent technologies, USA). Samples were injected in splitless mode at 320 °C. Compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness; J&W Scientific, USA). The carrier gas was helium flowing at a rate of 1.3 ml/min. The oven temperature was set at 90 °C for 1 minute, then increased at a rate of 5 °C/min to 320 °C and held for 23 minutes. In addition to scan mode, a selected ion monitoring (SIM) mode was used to further detect miliacin (olean-18-en-3β-ol methyl ether; *m/z* 189, 204, 231, 425, 440) [@Heron2016]. The oven temperature was set at 50 °C for 1 min, then raised at a rate of 10°C/min to 310 °C and held for 15 min. An authentic standard of miliacin was also run to monitor the retention time.

The GC-FID analysis of FAMEs was carried out using an Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode (Agilent technologies, USA). The carrier gas was helium flowing at a rate of 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was set at 90°C for 1.3 minute, then raised at a rate of 5°C/min to 320°C and held for 22 minutes. The GC-C-IRMS analyses of FAMEs of palmitic (C~16:0~) and stearic (C~18:0~) were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface (Thermo Fisher Scientific, USA) in splitless mode. Separation was carried out using the same model of column for GC-MS. Helium was used as a gas carrier with a constant flow rate of 1.24 ml/minute. The oven temperature was set at 90 °C for 1.5 minutes, then raised at a rate of 5 °C/min to 320 °C and held for 18 minutes.

@@

# add sample name
meth_all <-
  meths %>%
  mutate(sample = paste0(parse_number(sample),
                         str_extract(sample, pattern = "_00[0-9]"))) %>%
  mutate(name = case_when(sample %in% c("210729_001") ~ "CDG-062",
                          sample %in% c("210729_002") ~ "SYG-3",
                          sample %in% c("210729_003") ~ "SYG-TN13-E22-2#1",
                          sample %in% c("210729_004") ~ "SYG-TN13-E22-1",
                          sample %in% c("210729_005") ~ "SYG-TN13-E22-2#2",
                          sample %in% c("210729_006") ~ "SYG-TN13-E23-3")) %>%
  mutate(name = factor(name, levels = c("SYG-TN13-E22-1",
                                        "SYG-TN13-E22-2#1",
                                        "SYG-TN13-E22-2#2",
                                        "SYG-TN13-E23-3",
                                        "SYG-3",
                                        "CDG-062")))
```

```{r FA-composition-tidy, cache= TRUE, cache.extra=tools::md5sum(here::here("analysis/data/raw_data/210824_FAMES_pot_LW_publication.xlsx"))}
quant_FA <-
  readxl::read_excel(here::here("analysis", "data", "raw_data", "210824_FAMES_pot_LW_pub.xlsx"), 
                     skip = 2, col_names = TRUE)

quant_FA_number <-
  quant_FA %>%
  pivot_longer(!number&!sample&!`g extracted`&!`%reco`,
               names_to = "FA", 
               values_to = "weight") %>% 
  filter(str_starts(FA, "ng") & !sample =="") %>% 
  mutate(FA = str_remove(FA, "ng "),
         weight = round(weight/`%reco`/`g extracted`/1000, 2)) %>% 
  select(-number, -`%reco`, -`g extracted`) %>% 
  pivot_wider(names_from = sample, 
              values_from = weight) %>% 
  filter(!FA %in%c("C34:0", "C21:0"))

FA_sum_min <- sum(quant_FA_number$"SYG-TN13-E22-2#1", na.rm = T)
FA_sum_max <- sum(quant_FA_number$"SYG-TN13-E22-2#2")
```

The total lipid concentrations of potsherds at Guijiabao and Gaoshan range from `r FA_sum_min` to `r FA_sum_max` μg/g (Table \@ref(tab:FA-composition-table)). Lipid yields from all potsherds are above interpretable concentrations (5 μg/g) [@Evershed2008experimental], suggesting those ceramics should be associated with food processing. The main compounds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid) and C~18:0~ (stearic acid), commonly derived from animal fats and plant oils (Figure \@ref(fig:chromatograms-plots-four)). Unsaturated fatty acid, C~18:1~, was detected in all samples and dominant in three, but we are unable to distinguish the C~18:1~ types based on current methods. The most interesting finding is that milliacin, the compound specific to broomcorn millet [@Heron2016; @Bossard2013; @Ganzarolli2018], was identified in one Neolithic sample (SYG-TN13-E22-2#2) from Guijiabao (Figure \@ref(fig:chromatograms-specific-std)). This sample has a wider range of fatty acids from C~12:0~ to C~30:0~ (`r FA_sum_max` μg/g), where long-chain fatty acids could originate from plant sources [@Post1996; @Copley2005]. Overall, the dominance of C~16:0~ and C~18:0~ in most samples suggests that the lipids extracted from our samples were greatly degraded.

```{r FA-composition-table, cache= TRUE}
quant_FA_tidy <-
  quant_FA %>%
  rename("ng total" = "total FAs ug/g") %>%
  pivot_longer(!number&!sample&!`g extracted`&!`%reco`,
               names_to = "FA", 
               values_to = "weight") %>% 
  filter(str_starts(FA, "ng") & !sample =="") %>% 
  filter(!FA == "ng C34:0" & !FA == "ng C21:0") %>% 
  mutate(FA = str_remove(FA, "ng")) %>% 
  mutate(weight = ifelse(str_detect(FA, "C"), 
                         round(weight/`%reco`/`g extracted`/1000, 2), weight)) %>% 
  select(-number, -`%reco`, -`g extracted`) %>% 
  mutate(weight = round(weight, 2),
         weight = ifelse(is.na(weight), "-", weight),
         weight = ifelse(weight == 0, "-", weight)) %>% 
  pivot_wider(names_from = sample, 
              values_from = weight) %>% 
  select("FA", "SYG-TN13-E22-1", "SYG-TN13-E22-2#1", "SYG-TN13-E22-2#2",
         "SYG-TN13-E23-3", "SYG-3", "CDG-062") %>% 
  mutate(FA= str_replace(FA, "C(.*)", "C~\\1~"))

```{r highlight-compounds, cache=TRUE}
# round all peaks in order
all_peaks <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE)
